Lab 2 Final Report: COVID-19 Case Rate vs Population Demographics and Mask Policy

### 2020-12-08 Jackson_Liu_Temlock_Zhang_Lab2_Draft_Report.Rmd
Made grammatical and sentence structure edits across the entire document. Added some qualifying language around the poverty and unemployment rate data, and updated references of sections. Reworded variable effects to avoid causal implications

The research question aims to analyze the relationship between the COVID-19 case rate per 100,000 residents within each state and the population demographics features, as well as the policy decisions that were or were not put into place in order to combat the rise of COVID-19 cases. In this sense, the aim is to examine how dependent the COVID-19 case rate is on that of which cannot be controlled (i.e., population demographics), as well as that of which can be controlled to a certain degree (i.e., implementation of policies to attempt to combat case rate) by each state. With this information, one could suggest whether or not the proliferation of COVID-19 within a state seems to be related to either or both controllable and uncontrollable factors and how they differ. The modeling goal of this research question will be one of description, and will be broken up into three phases of investigation.

The first and primary phase of investigation involves how the COVID-19 case rate is related to key population demographics. For this, the key variables are COVID-19 case rate per 100,000 residents and the percentage population distribution of seniors (defined as 65 years old or older in the original dataset) within a state. The case rate per 100,000 was selected as the key output/dependent variable as it provides a standardized measure of the spread of COVID-19 across the states, and already takes into account the absolute population of the states. The senior percentage was selected as the key input/independent variable given that guidance had been released by the Centers for Disease Control and Prevention (CDC) that seniors belong to the age category for those at higher risk of COVID-19, and thus are a well studied and documented group. This is likely due to the both the fact that seniors in this age group are more likely or had greater access to be tested given their categorization and the fact that they would have a greater likelihood of exhibiting detectable COVID-19 related symptoms that would prompt them to get tested. An additional rationale for measuring the senior percentage is that they may also be more likely to contract COVID-19 at lower viral loads, making them more susceptible to the virus. Conversely, this may result in seniors being more wary of the threat of COVID-19, and taking additional precautions to prevent infection relative to other age categories, such as limiting social interactions and taking more preventative measures in terms of hygiene. These key variables will provide an initial understanding of the relationship between the spread of the virus and population demographics.

Following this, in the second phase of investigation the analysis considers other variables of state-level population demographic features that are considered factors of susceptibility, namely the rate of poverty (defined as the percent of individuals living under the federal poverty line in 2018 in the original dataset) and the rate of unemployment (defined as the percent unemployment in 2018 in the original dataset). Those living below the poverty line may have less access/are unable to afford to preventative controls such as sanitation products and masks that help prevent the spread of the virus, while those who are designated as unemployed may not have the flexibility of sheltering at home as well as access to the aforementioned preventative controls. Alternatively, the likelihood of reduced mobility for those living under the poverty line and/or who are unemployed may also play a role in the relationship with case rate. As mentioned, these variables operationalize population metrics that may lead to greater case rates, and are already standardized as rates to account for varying absolute populations across states.

The tertiary and final phase of investigation includes a variable that measures the implementation of policy as a response to COVID-19, and to understand the added effect of this variable in conjunction with the first phase of investigation. Specifically, there is a focus on a policy that mandates the wearing of masks within the state. Through this, the aim is to analyze whether the implementation of a mask-related policy is related to case rate, as well as the strength of the association compared to population metrics. To operationalize this policy and simplify the measurements, the models only consider whether or not this policy was implemented through transformed indicator variables (1 = implemented, 0 = not implemented). As a result, factors of when and for how long the policy was implemented will be lost. Although this loss of information may fail to capture the impact of length of time of a policy on case rate (the policy may take time to be adopted/show meaningful efficacy), as there is no time-series data for case rate included in the data set, it was adjudged to be incongruent with the analysis.

@@

Based on the work of model 1, two extra state demographic features regarding poverty rate (percentage living under the federal poverty line in 2018) and unemployment rate (percentage of unemployed in 2018) would be included in the analysis of model 2. Although the increasing unemployment rate during the pandemic has been widely covered by the media, the potential pandemic causing poverty or unemployment will not be investigated here, according to the designed research question in this work.

@@

```{r}
ggplot(data = df,
  mapping = aes(x= poverty_per_100k))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(6000,21200,800)) +
  labs(title =  "Histogram of Poverty Rate", 
       x = "Number of people living under poverty line per 100,000", y = 'Count')+
  scale_x_continuous(breaks=seq(6000, 21200, 1600))
```
As shown in the above distribution, there is a fairly normal distribution in the poverty number per 100,000 across the states. Additionally, there is no outlier observed within the distribution, given that the range of rates fall between 7,600 and 19,700. Therefore, there is no further transformation in need for this variable in the following analysis process for both model 2 and model 3.

@@

```{r}
ggplot(data = df,
  mapping = aes(x= unemployed_per_100k))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(2000,8000,500)) +
  labs(title =  "Histogram of Unemployment Rate", 
       x = "Number of unemployed per 100,000", y = 'Count')+
  scale_x_continuous(breaks=seq(2000, 8000, 1000))
```
According to the histogram above, the distribution of unemployed number per 100,000 is fairly normal distribution. Overall, it is not highly skewed or heavily tailed. Therefore, there is no further transformation needed for this variable in the following analysis process for both model 2 and model 3.

@@

Compared to model 1, the relationship investigated here involves a higher dimensional space. It would be challenging to observe notable variation from variables distribution plots as below.

```{r message=FALSE}
case_poverty <- df %>% 
  ggplot(aes(poverty_per_100k, case_rate_100k, color = senior_per_100k)) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'COVID-19 case rate vs poverty rate', 
    x = 'Number of living under poverty line per 100,000',
    y = 'Case rate per 100,000', 
    color = 'Seniors per 100k'
  )

case_unemployed <- df %>% 
  ggplot(aes(unemployed_per_100k, case_rate_100k, color = senior_per_100k)) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'COVID-19 case rate vs unemployment rate', 
    x = 'Number of unemployed per 100,000', 
    y = 'Case rate per 100,000',
    color = 'Seniors per 100k'
  )

@@

Based on the above data exploratory result, the following equation is used to create a regression model to determine the relationship between the case rate and demographic features including the senior rate, the rate of poverty and the rate of unemployment.

@@

According to the assessment of homoskedasticity of model 2, which would be discussed in the following paragraph (Model 2 Limitations - iv.Homoskedastic Errors), it is plausible to apply the classical standard errors in the t-test of coefficients here.

@@

The coefficient of the senior rate (senior_per_100k) is -0.230970 and statistically significant, which implies that for every one thousand additional seniors per 100,000, there is a decrease of 231 COVID-19 cases per 100,000, holding other demographic features as constant. As discussed in mode1 1 section, although seniors are susceptible group, less mobility and social interaction with cautious self protection may play an important role in this pandemic situation.

In addition, the coefficient of the poverty rate (poverty_per_100k) (0.231750) indicates that for every one thousand additional people living under poverty line per 100,000, there is a increase of 232 COVID-19 cases per 100,000, holding other demographic features as constant. People living under the poverty line have limited access to medical resources. Moreover, most of them could only afford to live in the communities with poor hygiene conditions.

Finally, the coefficient of the unemployment rate (unemployed_per_100k) is statistically significant and negative (-0.486191), which implies that for every one thousand additional unemployed, there is a decrease of 486 COVID-19 case per 100,000, holding other demographic features as constant. Although the unemployment leads to less medical insurance coverage and impact on their income, the unemployed people may have more flexibility to obey stay-at-home orders. In addition, there is less exposure probability when there is no commuting or in-place working required for the unemployed.

@@

#### ii. Linear Conditional Expectation
To assess the linear conditional expectation assumption in higher-dimensional space, the plots of predicted versus residuals of the model 2 is investigated as follows.

@@

Overall, the residuals maintain around zero across the predicted range. There is no assumption violation phenomenon that can be observed here.

@@

Since there are multiple covariates included in this model compared to the model 1, there is more concern about the no perfect collinearity assumption here. First, the list of variable coefficients shows that no variables were dropped, which means there is no perfect collinearity detected by the R function.

@@

#### iv. Homoskedastic Errors
To evaluate the homoskedastic errors assumption, the square root of the residuals is plotted against the fitted values of model 2 (scale-location plot). Although, the curve is not perfectly flat, there is no obvious variance of the errors either. For reference, the data points marked in the plot represent Florida (10), New York (33) and North Dakota (35).

@@

#### v. Normally Distributed Errors
To access the assumption of normality of error distribution, both the histogram and the Q-Q plot of the residuals is shown as follow.

@@

For model 3, policy related variables will be introduced for the analytics. Policy variable "Mandate face mask use by all individuals in public spaces" (renamed as mask_public) is selected as control variable to predict the case rate. Given it has been proved that mask is effective in preventing the transmission of disease, it is expected that the mask policy mandate in public would have an effect in reducing the case rate.

There are other variables related to mask such as "No legal enforcement of face mask mandate" and "Mandate face mask use by employees in public-facing businesses", it is assume these variables have a lower impact than the policy that is enforced to public. Also, it is observed that if a state enforce public mask policy, they will by default enforce mask policy for employees in public-facing business. As a result, mask policy for all individuals in public space can be used to represent mask-related policies.

It is worth to note that, the mask_public is in date format, which has value of 0 or a actual date when the policy is enforced. According to the documentation, 0 represents "the absence of an order or directive", which can be interpreted as the policy is not enforced by the state explicitly. For linear regression, the date values in mask_public column are transformed as value 1, so that   whether a state has any public mask policy or not can be distinguished. It is also noted that by transforming the variable, some important information will be lost because the actual date (early or late) to enforce the policy can also have an impact on case rate. However, it is difficult to measure the effect of the timing of implementation across different states. The enforcement of public mask policy will vary across states depending on other state related features such as political factors that are not included.

Firstly, the mask policy (mask_public_bool) variable is created from the original variable (mask_public), with 1 representing that the state has a pubic mask policy, and 0 means there's no explicit public mask policy from the state. The summary shows that there are 35 states that enforced public mask mandate, and 16 states that had no public mask policy at the time the dataset was compiled. The ratio looks reasonable for the analysis considering sample size of 51 as there is decent number of samples for each state policy type.

@@

cat('Count of states that enforced mask mandate in public: ', 
    length(df$mask_public_bool[df$mask_public_bool==1]))

cat('\nCount of states did not enforce mask mandate in public: ', 
    length(df$mask_public_bool[df$mask_public_bool==0]))
```

@@

```{r}
df %>% 
  ggplot(aes(x = mask_public_bool, y = case_rate_100k)) + 
  geom_point(aes(size=5, alpha=0.2), show.legend = FALSE) + 
  labs(
    title = 'Case Rate vs Mask Public Mandate', 
    x = 'Mask Mandate in Public', 
    y = 'Case Rate per 100,000')  +
  scale_x_continuous(breaks = seq(0,1,1), labels = seq(0,1,1))
```

@@

As can be seen from the histogram, although most of the population density clusters in the 0 to 1500 range, there are some grouping of outliers that are very far from this concentration. When an analysis is performed, it can be seen that there is only one data sample that is the outlier, which is D.C. with a value of 11,496. This is given due to the fact that D.C. is a district that solely consists of a large city, as mentioned as a possibility in the introduction. Given that this causes the data to be skewed, the logarithm is taken to scale the variable. Alternatively, the data point could have been dropped from the sample, but given the already small sample size, it was determined that a better approach would be to keep it within the sample. Once this transformation was performed, it is shown that there is a relatively normal distribution of population densities (see figure below).

@@

Next, the regression model for model 3 is created based on the existing work from model 2. Two additional variables discussed in Model 3 will be introduced as the maximalist approach. The first variable is the "mask_public_bool" (whether the state enforced public mask mandate or not) that represents the policy. The second variable introduced is "population_density" (the variable is transformed by the natural logarithm as discussed earlier) which is considered an important factor related to the spread of disease.

The regression coefficient shows that Mask_public_bool variable is highly significant. And the result suggests that public mask mandate policy has positive effect in bringing the case number down. Specifically, the COVID-19 case rate can be reduced by 945 out of every 100k people if the state enforces public mask mandate keeping all other variables the same.

The population density variable however is not significant in the regression even it is considered as one of the major factors to the spread of disease. According to the regression, there is a failure to reject the null hypothesis that population density is not related to the COVID-19 case rate. There could be multiple factors causing the population density to be insignificant. For example, the population density is calculated by population divided by total square miles. As a result the numbers can be diluted for states that have major cities with high population density, but very little urbanization in the majority part of the state suburb area. In addition, as stated earlier in subsection 2-3-1 of this report, Washington, D.C. is considered as a state however the demography structure and geographical related characteristics are different from other states. And due to the small sample size, individual variables that exhibit unique characteristics could lead to a certain degree of bias.

It was decided that the "population_density" variable should be kept in model 3 even it is insignificant for the following reasons. First, adding population density variable has little effect on the change of coefficients of control variables in model 2, which helps to demonstrate the robustness of the results. Second, there is no evidence that the variable effects the validation of CLM assumptions compared to model 2. Finally, the variable offers meaningful information that is relevant to the main research question.

@@

The residual versus fitted plot shows an almost flat line that is close to zero, with the tails a bit skewed. This implies that there are no strong linear relationships that are not captured beyond the current control variables.

@@

First, the list of variable coefficients shows that no variables were dropped, which means there is no perfect collinearity detected by the R function.

@@

Next, the variance inflation function was run in order to check for strong collinearity. Based on the variance inflation factors check, all the variables have variance inflation factors less than 4. As a result, there's no strong evidence of collinearity concern within the model.

@@

According to the scale-location plot, the standardized residual falls into a reasonable range and the curve looks smooth without strong deviations. There is no significant evidence that the model violates the homoskedasticity assumption. For reference, the marked data points are New Hampshire (30), New York (33), and North Dakota (35).

@@

```
Additionally, a Breusch-Pagan test is run to check the level of heteroskedasticity. The test result shows it fails to reject the null hypothesis, which indicates that there is no strong evidence of heteroskedasticity in model 3.

@@

To access the assumption of normality of error distribution, both the histogram and the Q-Q plot of the residuals is shown as follow.

@@

There is a fairly normal distribution that can be seen from the histogram above. From the Q-Q plot, the points generally stay close to the normal line, with the tails exhibiting a certain degree of skew. However, in general, it implies no deviation from normality. Overall, the normally distributed errors assumption is satisfied for model 3.

@@

The most general requirement for all the linear models is that data points be independent and identically distributed (IID) which they are built off of. While for the most part the events of one state will primarily affect that single state, travel by individuals spreading the virus across state lines is an inevitable occurrence. This may cause the spread of a virus within one state to be influenced by those around it, or generally by its position in the country. In addition, all states are involved in the same market economy system of the country to varying degrees, even though there are economic performance divisions across the states. Likewise, certain regions of the country may also have similar demographics due to a shared history across general geographic regions. While this is acknowledged to occur, the effect is likely minor compared to the policy data collected state by state which would affect the spread of the virus. Therefore it is relatively safe for this assumption to be met.

The identical distribution of states as individual samples in the model is also questionable. Similar to how connections or commonalities between states make them less than truly independent, certain factors would also influence the distribution so that some states are representative of a different underlying distribution than one another. For example, if one geographic region as a whole has a factor which influences the spread of the virus differently than another, multiple underlying distributions may be present in the sample. If still a new geographic region with many states was added to the sample, and the measures of central tendency of the data continued to evolve, there would be a further indication that the underlying distribution is not identical. Without a possibility of refining the data sample however, as states would be immutable in this regard, the regression analysis proceeds with the recognition of possible violations of the IID assumption.

In examining the results of various diagnostics across different models, generally the CLM assumptions are better met as the model continues to develop. In the first model, deviations away from certain assumptions are the strongest. This is seen mostly with regards to the linear conditional expectation assumption, where the residuals against the fitted values have a noticeable inverted "V" shape as opposed to being constant around zero. In the second model, the same diagnostic shows that the assumption is better met. This is further improved upon in the third model, where local variations in the linear conditional expectation diagnostic plot are generally smoother and more centered around zero. The assumption with the greatest exception to this is that of no high collinearity within the models, as each coefficient in each model was shown to have VIF values not near the cut off value of four.

However, when adding more variables in the third model, other assumptions begin to be less well met. For example, the normality of the error term in the model is less well fit than it was in the earlier models. It can be seen in the third model's Q-Q plot how at fitted values towards the edges of the model's range, there exists the greatest deviation away from normality. This assumption is generally less important than that of the linear conditional expectation, however, which is a fundamental part of CLM. Therefore, the third model can still be considered as generally improving the overall analysis.

In addition, with more variables in the third model, outliers appear to have more of an impact than previously. This is likely due to the limited number of data points the model is based off of, an entire state only constitutes an individual datapoint. With less degrees of freedom in the third model, certain points can be seen to more closely approach the boundaries of the Cook's length plot, although they are still not extreme enough to fully be considered an outlier. Increasing the resolution of the data, such as examining COVID-19 cases at the county or city/town level, would help to reduce this by increasing the overall degrees of freedom.

@@

Omitted variables from the model may affect both the internal specifications of included variables as well as the general applicability of the model to explain the output variable. For this analysis, the omitted variables mentioned are hypothetical. Therefore, without data to include in an follow up analysis with the omitted variable, only the direction of the bias may be estimated rather than the size effect as well.

>>>>

The research question aims to analyze the relationship between the COVID-19 case rate per 100,000 residents within each state (recorded between January 21, 2020 and October 30, 2020) and the population demographics features, as well as the policy decisions that were or were not put into place in order to combat the rise of COVID-19 cases. In this sense, the aim is to examine how dependent the COVID-19 case rate is on that of which cannot be controlled (i.e., population demographics), as well as that of which can be controlled to a certain degree (i.e., implementation of policies to attempt to combat case rate) by each state. With this information, one could suggest whether or not the proliferation of COVID-19 within a state seems to be related to either or both controllable and uncontrollable factors, and how they differ. The modeling goal of this research question will be one of description, and will be broken up into three phases of investigation.

The first and primary phase of investigation involves how the COVID-19 case rate is related to key population demographics. For this, the key variables are COVID-19 case rate per 100,000 residents and the percentage population distribution of seniors (defined as 65 years old or older in the original dataset) within a state. The case rate per 100,000 was selected as the key output/dependent variable as it provides a standardized measure of the spread of COVID-19 across the states, and already takes into account the absolute population of the states. The senior percentage was selected as the key input/independent variable given that guidance had been released by the Centers for Disease Control and Prevention (CDC) that seniors belong to the age category for those at higher risk of COVID-19, and thus are a well studied and documented group. This is likely due to both the fact that seniors in this age group are more likely or had greater access to be tested given their categorization, and the fact that they would have a greater likelihood of exhibiting detectable COVID-19 related symptoms that would prompt them to get tested. An additional rationale for measuring the senior percentage is that they may also be more likely to contract COVID-19 at lower viral loads, making them more susceptible to the virus. Conversely, this may result in seniors being more wary of the threat of COVID-19, and taking additional precautions to prevent infection relative to other age categories, such as limiting social interactions and taking more preventative measures in terms of hygiene. These key variables will provide an initial understanding of the relationship between the spread of the virus and population demographics.

Following this, in the second phase of investigation the analysis considers other variables of state-level population demographic features that are considered factors of susceptibility. Specifically, these features are the rate of poverty (defined as the percent of individuals living under the federal poverty line in 2018 in the original dataset) and the rate of unemployment (defined as the percent unemployment in 2018 in the original dataset). Those living below the poverty line may have less access/are unable to afford preventative controls such as sanitation products and masks that help prevent the spread of the virus, while those who are designated as unemployed may not have the flexibility of sheltering at home nor have access to the aforementioned preventative controls. Alternatively, the likelihood of reduced mobility for those living under the poverty line and/or those who are unemployed may also play a role in the relationship with case rate. As mentioned, these variables operationalize population metrics that may lead to greater case rates, and are already standardized as rates to account for varying absolute populations across states. It should be qualified that the data for poverty rate and unemployment rate are from 2018, and therefore may not provide a fully accurate reflection of the respective rates at the time that the case rate data was recorded.

The tertiary and final phase of investigation includes a variable that measures the implementation of policy as a response to COVID-19, and to understand the added effect of this variable in conjunction with the other phases of investigation. Specifically, there is a focus on a policy that mandates the wearing of masks in public spaces within the state. Through this, the aim is to analyze whether the implementation of a mask-related policy is related to case rate, as well as the strength of the association relative to that of population metrics. To operationalize this policy and simplify the measurements, the model will only consider whether or not this policy was implemented through a transformed indicator variable (1 = implemented, 0 = not implemented). As a result, factors of when and for how long the policy was implemented will be lost. Although this loss of information may fail to capture the impact of length of time of a policy on case rate (the policy may take time to be adopted/show meaningful efficacy), as there is no time-series data for case rate included in the data set, it was adjudged to be incongruent with the analysis.

@@

Building upon the work of model 1, two extra state demographic features regarding poverty rate (percentage living under the federal poverty line in 2018) and unemployment rate (percentage of unemployed in 2018) are included in the analysis of model 2. Although the increasing unemployment rate during the pandemic has been widely covered by the media, the potential pandemic causing poverty or unemployment will not be investigated here, according to the designed research question in this work.

@@

```{r}
ggplot(data = df,
  mapping = aes(x= poverty_per_100k))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(6000,21200,800)) +
  labs(title =  "Histogram of Poverty Rate", 
       x = "Number of people living under poverty line per 100,000", y = 'Count')+
  scale_x_continuous(breaks=seq(6000, 21200, 1600))
```
As shown in the above distribution, there is a fairly normal distribution in the poverty number per 100,000 across the states. Additionally, there is no outlier observed within the distribution, given that the range of rates fall between 7,600 and 19,700. Therefore, there is no further transformation needed for this variable in the following analysis process for both model 2 and model 3.

@@

```{r}
ggplot(data = df,
  mapping = aes(x= unemployed_per_100k))+
  geom_histogram(fill = 'skyblue', color = 'grey30', breaks = seq(2000,8000,500)) +
  labs(title =  "Histogram of Unemployment Rate", 
       x = "Number of unemployed per 100,000", y = 'Count')+
  scale_x_continuous(breaks=seq(2000, 8000, 1000))
```
According to the histogram above, the distribution of the unemployed number per 100,000 is fairly normal. Overall, it is not highly skewed or heavily tailed. Therefore, there is no further transformation needed for this variable in the following analysis process for both model 2 and model 3.

@@

Compared to model 1, the relationship investigated here involves a higher dimensional space. It would be challenging to observe notable variation from the variable distribution plots as below.

```{r message=FALSE}
case_poverty <- df %>% 
  ggplot(aes(poverty_per_100k, case_rate_100k, color = senior_per_100k)) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'COVID-19 case rate vs poverty rate', 
    x = 'Number of people living under poverty line per 100,000',
    y = 'Case rate per 100,000', 
    color = 'Seniors per 100,000'
  )

case_unemployed <- df %>% 
  ggplot(aes(unemployed_per_100k, case_rate_100k, color = senior_per_100k)) + 
  geom_point() + 
  geom_smooth(se=FALSE)+
  labs(
    title = 'COVID-19 case rate vs unemployment rate', 
    x = 'Number of unemployed per 100,000', 
    y = 'Case rate per 100,000',
    color = 'Seniors per 100,000'
  )

@@

Based on the results of the data exploration above, the following equation is used to create a regression model to determine the relationship between the case rate and demographic features including the senior rate, the rate of poverty and the rate of unemployment.

@@

According to the assessment of homoskedasticity of model 2, which would be discussed in the following subsection (Model 2 Limitations - iv.Homoskedastic Errors), it is plausible to apply the classical standard errors in the t-test of coefficients here.

@@

The coefficient of the senior rate (senior_per_100k) is -0.230970 and is statistically significant, which implies that for every one thousand additional seniors per 100,000, there is a decrease of 231 COVID-19 cases per 100,000, holding other demographic features constant. As discussed in the subsection 2-1, although seniors are a susceptible group, their reduced mobility and social interactions with cautious self protection may play an important role in reducing the number of cases among seniors.

In addition, the coefficient of the poverty rate (poverty_per_100k) (0.231750) indicates that for every one thousand additional people living under the poverty line per 100,000, there is an increase of 232 COVID-19 cases per 100,000, holding other demographic features constant. People living under the poverty line may have limited access to medical resources. Moreover, they may not be able to afford to live in communities with appropriate sanitary conditions.

Finally, the coefficient of the unemployment rate (unemployed_per_100k) is statistically significant and negative (-0.486191), which implies that for every one thousand additional unemployed per 100,000, there is a decrease of 486 COVID-19 case per 100,000, holding other demographic features constant. Although unemployment leads to less medical insurance coverage and impacts income, the unemployed may have more flexibility to obey stay-at-home orders. In addition, there is less exposure probability given the unemployed do not commute to work or work in-person.

@@

#### ii. Linear Conditional Expectation
To assess the linear conditional expectation assumption in a higher-dimensional space, the plot of the predicted values versus the residuals of model 2 is investigated as follows.

@@

Overall, the residuals remain around zero across the predicted range. There is no assumption violation phenomenon that can be observed here.

@@

Since there are multiple covariates included in this model compared to the model 1, there is concern about the no perfect collinearity assumption. First, the list of variable coefficients show that no variables were dropped, which means there was no perfect collinearity detected by the R function.

@@

#### iv. Homoskedastic Errors
To evaluate the homoskedastic errors assumption, the square root of the residuals is plotted against the fitted values of model 2 (scale-location plot). Although, the curve is not perfectly flat, there is no obvious variance in the errors either. For reference, the data points marked in the plot represent Florida (10), New York (33) and North Dakota (35).

@@

#### v. Normally Distributed Errors
To access the assumption of normality of error distribution, both the histogram and the Q-Q plot of the residuals are shown as follows.

@@

For model 3, policy related variables will be introduced for the analytics. The policy variable "Mandate face mask use by all individuals in public spaces" (renamed as mask_public) is selected as control variable to predict the case rate. Given it has been proven in external research that masks are effective in preventing the transmission of COVID-19, it is expected that the mask policy mandate in public would have an effect in reducing the case rate.

There are other variables related to masks such as "No legal enforcement of face mask mandate" and "Mandate face mask use by employees in public-facing businesses". It is assumed these variables have a lower impact than the policy that is enforced to those in public. Also, it is observed that if a state enforces a public mask policy, they will also enforce a mask policy for employees in public-facing businesses. As a result, mask policy for all individuals in public spaces can be used to represent mask-related policies.

It is worth noting that the mask_public variable is in date format, which has either a value of 0 or an actual date when the policy was enforced. According to the documentation, 0 represents "the absence of an order or directive", which can be interpreted as the policy is not enforced by the state explicitly. For the linear regression, the date values in the mask_public variable are transformed into a value of 1, so that whether or not a state has a public mask policy can be distinguished. It is also noted that by transforming the variable, some important information will be lost because the actual date (early or late) of enforcement for the policy can also have an impact on case rate. However, it is difficult to measure the effect of the timing of implementation across different states. Additionally, the enforcement of a public mask policy will vary across states depending on other state related features such as political factors that are not included within this analysis.

Firstly, the mask policy (mask_public_bool) variable is created from the original variable (mask_public), with 1 representing that the state has a pubic mask policy, and 0 meaning there is no explicit public mask policy in the state. The output below shows that there are 35 states that enforced a public mask mandate and 16 states that did not at the time the dataset was compiled. This ratio looks reasonable enough for analysis, considering that there is a decent number of samples within each group from the original sample size of 51.

@@

cat('Count of states that enforced a mask mandate in public spaces: ', 
    length(df$mask_public_bool[df$mask_public_bool==1]))

cat('\nCount of states did not enforce a mask mandate in public spaces: ', 
    length(df$mask_public_bool[df$mask_public_bool==0]))
```

@@

```{r, message=FALSE, warning=FALSE}
df %>% 
  ggplot(aes(x = mask_public_bool, y = case_rate_100k)) + 
  geom_point(aes(size=5, alpha=0.2), show.legend = FALSE) + 
  labs(
    title = 'Case Rate vs Mask Public Mandate', 
    x = 'Mask Mandate in Public', 
    y = 'Case Rate per 100,000')  +
  scale_x_continuous(breaks = seq(0,1,1), labels = seq(0,1,1))

df %>%
  ggplot(aes(x=mask_public_bool, y=case_rate_100k))+
  geom_jitter(shape=21, size=6, aes( fill = factor(mask_public_bool)), width=0.1)+
  theme(legend.position = 'none') +
  labs(
    title = 'Case Rate vs Mask Public Mandate',
    x = 'Mask Mandate in Public',
    y = 'Case Rate per 100,000',
    color = 'Case Rate'
  ) +
  scale_x_continuous(breaks = seq(0,1,1), labels = seq(0,1,1))

df %>%
  ggplot(aes(case_rate_100k))+
  geom_histogram(aes(y = ..density.., fill = factor(mask_public_bool)), position = 'identity', alpha = 0.5, color = 'white')+
  stat_density(geom = 'line', size=1, position = 'identity', aes(color = factor(mask_public_bool)))+
  scale_fill_brewer(palette = 'Dark2') +
  scale_color_brewer(palette = 'Dark2') +
  labs(
    title = 'Distribution of COVID-19 Case Rate',
    x = 'Covid-19 case rate per 100,000'
  ) +
  scale_fill_discrete(name='Mask public policy',labels=c('No', 'Yes'))+
  scale_color_discrete(name='Mask public policy',labels=c('No', 'Yes'))
```

@@

As can be seen from the histogram, although most of the population density clusters in the 0 to 1500 range, there is some grouping of outliers that are very far from this concentration. When an analysis is performed, it can be seen that there is only one data sample outlier, which is D.C. with a value of 11,496. This is given due to the fact that D.C. is a district that solely consists of a large city. Given that this causes the data to be skewed, the logarithm is taken to scale the variable. Alternatively, the data point could have been dropped from the sample, but given the already small sample size, it was determined that a better approach would be to keep it as a data point. Once this transformation was performed, it is shown that there is a relatively normal distribution of population densities (see figure below).

@@

Next, the regression model for model 3 is created based on the existing work from model 2. Two additional variables discussed in model 3 will be introduced given the maximalist approach for model 3. The first variable is the "mask_public_bool" (whether the state enforced a public mask mandate or not) that represents the policy. The second variable introduced is "population_density" (the variable is transformed by the natural logarithm as discussed earlier) which is considered an important factor related to the spread of disease.

The regression coefficient shows that mask_public_bool variable is highly significant. The result also suggests that the public mask mandate policy has a negative relationship with the case rate. Specifically, the COVID-19 case rate decreases by 945 out of every 100,000 people if the state enforces a public mask mandate, holding all else constant.

Conversely, the population density variable is not significant in the regression even though it is considered as a major factor in stymieing the spread of the virus. According to the regression, there is a failure to reject the null hypothesis that population density is not related to the COVID-19 case rate. There could be multiple factors causing the population density to be insignificant. For example, the population density is calculated by population divided by total square miles. As a result, the numbers can be diluted for states that have major cities with high population density, but very little urbanization in the majority of the rural area. In addition, as stated earlier in subsection 2-3-1 of this report, D.C. is considered as a state despite the fact that the demographic and geographical characteristics differ from other states. And due to the small sample size, individual variables such as this that exhibit unique characteristics could lead to a certain degree of bias.

It was decided that the "population_density" variable should be kept in model 3 despite its insignificance for several reasons. First, adding population density variable has little effect on the coefficients of the control variables in model 2, which helps to demonstrate the robustness of the results. Second, there is no evidence that the variable effects the validation of the CLM assumptions compared to those in model 2. Finally, the variable offers meaningful information that is relevant to the main research question.

@@

The residual versus fitted plot shows an almost flat line that is close to zero, with the tails slightly skewed. This implies that there are no strong linear relationships that are not captured beyond the current control variables.

@@

First, the list of variable coefficients shows that no variables were dropped, which means there was no perfect collinearity detected by the R function.

@@

Next, the variance inflation function was run in order to check for strong collinearity. Upon examination of the ouptut below, it can be seen that all the variables have variance inflation factors of less than 4. As a result, there is no strong evidence of collinearity within the model.

@@

According to the scale-location plot, the standardized residuals fall into a reasonable range and the curve looks smooth without strong deviations. There is no significant evidence that the model violates the homoskedasticity assumption. For reference, the marked data points are New Hampshire (30), New York (33), and North Dakota (35).

@@

Additionally, a Breusch-Pagan test is run to check the level of heteroskedasticity. The test result shows it fails to reject the null hypothesis, which indicates that there is no strong evidence of heteroskedasticity in model 3.

@@

To assess the assumption of normality of error distribution, both the histogram and the Q-Q plot of the residuals are shown as follows.

@@

There is a fairly normal distribution that can be seen from the histogram above. In the Q-Q plot, the points generally stay close to the normal line, with the tails exhibiting a certain degree of skew. However, in general, it implies no deviation from normality. Overall, the normally distributed errors assumption is satisfied for model 3.

@@

The most general requirement for all the linear models is that data points be independent and identically distributed (IID). While for the most part the events of one state will primarily affect that single state, travel by individuals spreading the virus across state lines is an inevitable occurrence. This may cause the spread of a virus within one state to be influenced by those around it, or generally by its geographical location within the country. In addition, all states are involved in the same market economy system to varying degrees, even though there are economic performance divisions across the states. Likewise, certain regions of the country may also have similar demographics due to a shared history across general geographic regions. While this is acknowledged to occur, the effect is likely minor compared to the policy data collected state by state which would affect the spread of the virus. Therefore it is relatively safe to reason that this assumption has been met.

The identical distribution of states as individual samples in the model is also questionable. Similar to how connections or commonalities between states make them less than truly independent, certain factors would also influence the distribution so that some states are representative of a different underlying distribution compared to others. For example, if one geographic region as a whole has a factor which influences the spread of the virus differently than another, multiple underlying distributions may be present in the sample. If still a new geographic region with many states was added to the sample, and the measures of central tendency of the data continued to evolve, there would be a further indication that the underlying distribution is not identical. Without any possibility of refining the data sample however, as states would be immutable in this regard, the regression analysis proceeds with the recognition of possible violations in the IID assumption.

In examining the results of various diagnostics across different models, the CLM assumptions are generally better met as variables continue to be added. In the first model, deviations away from certain assumptions are the strongest. This is seen mostly with regards to the linear conditional expectation assumption, where the residuals against the fitted values have a noticeable inverted "V" shape as opposed to being constant around zero. In the second model, the same diagnostic shows that the assumption is better met. This is further improved upon in the third model, where local variations in the linear conditional expectation diagnostic plot are generally smoother and more centered around zero. The assumption with the greatest exception to this is that of no high collinearity within the models, as each coefficient in each model was shown to have VIF values far below the cut off value of four.

However, when adding additional variables in the third model, other assumptions begin to be less well met. For example, the normality of the error term in the model is less well fit than it was in the earlier models. It can be seen in the third model's Q-Q plot that for fitted values at the edges of the model's range, there exists the greatest deviation away from normality. This assumption is generally less important than that of the linear conditional expectation, however, which is a fundamental part of the CLM. Therefore, the third model can still be considered as generally improving the overall analysis.

In addition, with more variables in the third model, outliers appear to have more of an impact than they did previously. This is likely due to the limited sample size within the models. With less degrees of freedom in the third model, certain points can be seen to more closely approaching the boundaries of the Cook's distance plot, although they are still not extreme enough to be considered outliers. Increasing the resolution of the data, such as examining COVID-19 cases at the county or city/town level, would help to reduce this by increasing the overall degrees of freedom.

@@

Omitted variables from the model may affect both the internal specifications of included variables as well as the general applicability of the model in explaining the output variable. For this analysis, the omitted variables mentioned are hypothetical. Therefore, without the data available to conduct a follow-up analysis on the omitted variables, only the direction of the bias may be estimated, as opposed to the actual effect size.
